Ecosystem size and complexity dictate riverine biodiversity

### 2021-01-26 manuscript.Rmd
Update manuscript.Rmd

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  html_notebook: default
  word_document: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesized that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity) that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two biogeographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

@@

Given the results of extensive sensitivity analysis (**Supplementaly Information**), we considered `r n_scenario` simulation scenarios, which comprised a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries [@chezikRiverNetworksDampen2017; @mooreEmergentStabilityLarge2015; @petersonModellingDendriticEcological2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@mooreSynchronizationPortfolioPerformance2010; @poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see **Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1000 time steps of metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data that span across two geographic regions: the Hokkaido island in Japan and the midwestern US (Midwest). These regions are located in comparable ranges of latitude but support distinct fish communities (**Figure 1**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Methods** and **Supplementary Information**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area), or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We combined this data set with geospatial information (watershed area, branching probability, annual mean temperature, annual cumulative precipitation, and fraction of forest) to examine potential influences of macro-scale drivers of species diversity. Using this data set, we developed global and region-specific models for each of the diversity measures ($\alpha$, $\beta$, and $\gamma$ diversity) to examine whether observed patterns are consistent across the two geographic regions. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity by including interaction terms with the dummy binary variable of the region (Hokkaido = 0; Midwest = 1). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (see **Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

Despite the substantial difference in fish fauna between the study regions, we found strong support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`): this result suggests the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two biogeographic regions. Furthermore, the observed patterns were in close agreement with our theoretical predictions (**Table 1 and Figure 4**). The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, and land use) (**Table 1**). Overall, our statistical analysis provides strong empirical evidence that ecosystem size and complexity jointly scale riverine biodiversity at the metacommunity level.

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor = `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a weaker or a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Direct (mark-recapture) and indirect observations (e.g., genotyping) recurrently revealed the restricted movement of stream fish, typically limited to several tens to hundreds of meters in distance[@comteFishDispersalFlowing2018; @radingerPatternsPredictorsFish2014; @rodriguezRestrictedMovementStream2002]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because the watersheds in the Midwest are severely altered by agriculture (median % agricultural land use: `r round(median(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(median(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans causes significant homogenization of in-stream habitat conditions among tributaries, theory predicts weakened effects of branching probability on $\gamma$ diversity. However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial turnover of species composition. In support of this interpretation, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes among tributaties[@bendaNetworkDynamicsHypothesis2004]. Although our analysis is correlative and cannot provide conclusive evidence, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], habitat fragmentation[@fukushimaModellingEffectsDams2007; @moritaEffectsHabitatFragmentation2002], and stream burial, may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and how to design a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020; @stollSmallImpoverishedRegional2013; @sundermannRiverRestorationSuccess2011]. While large reserves or restoration sites are undoubtedly important[@koningNetworkGrassrootsReserves2020], our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koningNetworkGrassrootsReserves2020]. Our findings, therefore, offer the ecological foundation for the network-based conservation strategies in rivers, broadening viable options towards successful conservation.

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity, ecosystem complexity - especially branching structure - has not received the attention it deserves. Ecosystems are inherently complex in their spatial structure, providing a physical template that creates a wide spectrum of niche opportunities for living organisms. Hence, our findings should be broadly applicable to many taxa and ecosystems. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

Two sources of variation characterize the long-term average of abiotic environment at each habitat patch: environmental variation at headwaters ($\sigma_h$) and local environmental noise ($\sigma_l$). We draw random values from a normal distribution with a mean of zero and SD of $\sigma_{h}$ and assigned them to headwater patches (i.e., the most upstream patches). These environmental values at headwaters propagate downstream through a spatial autoregressive process defined as:

@@

### Simulation
We set four landscape and eight ecological scenarios, resulting in a total of `r n_scenario` sets of parameter combinations. For landscape scenarios, we changed values of environmental variation at headwaters ($\sigma_h$) and the degree of local environmental noise ($\sigma_l$) separately to create contrasting patterns of habitat heterogeneity. Specifically, we used two values (0.01 and 1) for each of SD parameters. When $\sigma_h > \sigma_l$, branching exhibits important effects on habitat heterogeneity in the ecosystem because headwaters are the primary source of environmental variation. When $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity in the ecosystem.

@@

We compiled fish data at two biogeographic regions: the Hokkaido island, Japan and the Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish[@fukushimaHokkaidoFreshwaterFish2011; @fukushimaModellingEffectsDams2007], monitoring data at protected watersheds[@comteRivFishTIMEGlobalDatabase2021; @teruiMetapopulationStabilityBranching2018], and primary data collected by the authors[@miyazakiTemporalDynamicsFluvial2016; @teruiThreeEcologicalFactors2016], which collectively cover the entire Hokkaido island. For the Midwest, we assembled fish community data collected by Iowa Department of Natural Resources, Illinois Environmental Protection Agency and Illinois Department of Natural Resources, Minnesota Pollution Control Agency, and Wisconsin Department of Natural Resources. We detailed procedures for initial data selection in **Supplementary Information**.

@@

We collected data for watershed characteristics (watershed area [km^2^], branching probability, and mean elevation [m]), climatic variables (annual mean air temperature [degree celsius] and cumulative precipitation [mm]), and land use (fraction of forest, fraction of urban, fraction of agriculture, and dam density per watershed area [km^-2^]). We delineated watershed polygons and river lines (defined as $\ge$ 1 km^2^ in the watershed area) using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] with ArcMap 10.7 and SAGA 6.3.0 combined with the R package 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018]. Watershed area was estimated based on the delineated watershed polygons using functions in the R package 'sf'[@pebesmaSimpleFeaturesStandardized2018]. We estimated branching probability by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. We obtained mean elevations for each watershed polygon using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019]. We estimated the spatial averages of air temperature and cumulative precipitation at a 1-km resolution for each watershed, which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005] via the R package 'raster.' We calculated fraction of forest, urban, and agriculture for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020]. We calculated dam density per watershed area (km^-2^) based on the Global Reservoir and Dam Database, Version 1 (GRanDv1)[@lehnerGlobalReservoirDam2011; @lehnerHighResolutionMappingWorld2011].

@@

We developed linear models to examine the effects of macro-scale drivers on diversity metrics. Log-transformed diversity metrics ($\alpha$, $\beta$ and $\gamma$ diversity) at watershed $i$, $log_{10}~y_i$, are assumed to follow a normal distribution as $log_{10}~y_i \sim Normal(\mu_i, \sigma^2)$. The parameter $\mu_i$ was related to linear predictors, including log-transformed watershed area (ecosystem size; $log_{10}~WA$), log-branching probability (ecosystem complexity; $log_{10}~P_b$), region (coded as Hokkaido = 0, Midwest = 1; $region$), air temperature, precipitation, and fraction of forest land use (logit-transformed). Although our primary focus of this analysis is to assess the effects of watershed area and branching probability, we included climate and land use variables to control these effects statistically. Fractions of urban and agriculture land use were not included in the statistical models because these variables are strongly correlated with the fraction of forest land use (**Figure S19**). Also, we did not use mean elevation in the statistical models because it showed a moderate correlation with air temperature (**Figure S19**). Correlations for other combinations of explanatory variables were weak to moderate.

@@

We constructed two models for each of the diversity metrics: global and region-specific models. The global model assumes that the effects of watershed area and branching probability are consistent across biogeographic regions (i.e., no interactive effects of region with other linear predictors).

@@

# Acknowledgements
We acknowledge numerous scientists who have been involved in the collection of fish community data. We thank Michio Fukushima for providing information on the Hokkaido Freshwater Fish Database HFish.

@@

# Data and code availability
Data are provided by third parties and references are provided in the main text where appropriate. R functions for the generation of branching networks and the metacommunity simulation are provided as the R package 'mcbrnet.' (available at https://github.com/aterui/mcbrnet). Codes for simulations, statistical analysis, figures, and tables are available at https://github.com/aterui/public-proj_stream-diversity.

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  word_document: default
  html_notebook: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesized that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity) that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two biogeographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

@@

Given the results of extensive sensitivity analysis (**Supplementary Information**), we considered `r n_scenario` simulation scenarios, which comprised a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries [@chezikRiverNetworksDampen2017; @mooreEmergentStabilityLarge2015; @petersonModellingDendriticEcological2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@mooreSynchronizationPortfolioPerformance2010; @poffHomogenizationRegionalRiver2007; @teruiMetapopulationStabilityBranching2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see **Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1000 time steps of metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: the Hokkaido island in Japan and the midwestern US (Midwest). These regions are located in comparable latitude ranges but support distinct fish communities (**Figure 1**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Methods** and **Supplementary Information**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area), or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We combined this data set with geospatial information (watershed area, branching probability, annual mean temperature, annual cumulative precipitation, and the fraction of forest) to examine potential influences of macro-scale drivers of species diversity. Using this data set, we developed global and region-specific models for each of the diversity measures ($\alpha$, $\beta$, and $\gamma$ diversity) to examine whether observed patterns are consistent across the two geographic regions. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity by including interaction terms with the dummy binary variable of the region (Hokkaido = 0; Midwest = 1). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (see **Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

Despite the substantial difference in fish fauna between the study regions, we found strong support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`): this result suggests the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two biogeographic regions. Furthermore, the observed patterns were closely aligned with our theoretical predictions (**Table 1 and Figure 4**). The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, the fraction of forest, and dam density) (**Table 1**). Overall, our statistical analysis provides strong empirical evidence that ecosystem size and complexity jointly scale riverine biodiversity at the metacommunity level.

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor = `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a weaker or a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Direct (mark-recapture) and indirect observations (e.g., genotyping) recurrently revealed the restricted movement of stream fish, typically limited to several tens to hundreds of meters[@comteFishDispersalFlowing2018; @radingerPatternsPredictorsFish2014; @rodriguezRestrictedMovementStream2002]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because the watersheds in the Midwest are severely altered by agriculture (median % agricultural land use: `r round(median(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(median(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans causes significant homogenization of in-stream habitat conditions among tributaries, theory predicts a weakened effect of branching probability on $\gamma$ diversity. However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial turnover of species composition. In support of this interpretation, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes among tributaties[@bendaNetworkDynamicsHypothesis2004]. Although our analysis is correlative and cannot provide conclusive evidence, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], landscape modification[@finlayStreamSizeHuman2011; @moritaEffectsHabitatFragmentation2002], and stream burial[@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020; @stollSmallImpoverishedRegional2013; @sundermannRiverRestorationSuccess2011]. While large reserves or restoration sites are undoubtedly important[@koningNetworkGrassrootsReserves2020], our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koningNetworkGrassrootsReserves2020].

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity, ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial[@turnerLandscapeEcologyTheory2015] and aquatic ecosystems[@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of ecosystem structure. Our general framework of the dual dimensional approach, ecosystem size and complexity, should apply beyond branching ecosystems and broaden viable options towards successful conservation. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

Two sources of variation characterize the long-term average of the abiotic environment at each habitat patch: environmental variation at headwaters ($\sigma_h$) and local environmental noise ($\sigma_l$). We draw random values from a normal distribution with a mean of zero and SD of $\sigma_{h}$ and assigned them to headwater patches (i.e., the most upstream patches). These environmental values at headwaters propagate downstream through a spatial autoregressive process defined as:

@@

### Simulation
We set four landscape and eight ecological scenarios, resulting in a total of `r n_scenario` sets of parameter combinations. For landscape scenarios, we changed values of environmental variation at headwaters ($\sigma_h$) and the degree of local environmental noise ($\sigma_l$) separately to create contrasting patterns of habitat heterogeneity. Specifically, we used two values (0.01 and 1) for each of the SD parameters. When $\sigma_h > \sigma_l$, branching exhibits important effects on habitat heterogeneity in the ecosystem because headwaters are the primary source of environmental variation. When $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity in the ecosystem.

@@

We compiled fish data at two biogeographic regions: the Hokkaido island, Japan, and the Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish[@fukushimaHokkaidoFreshwaterFish2011; @fukushimaModellingEffectsDams2007], monitoring data at protected watersheds[@comteRivFishTIMEGlobalDatabase2021; @teruiMetapopulationStabilityBranching2018], and primary data collected by the authors[@miyazakiTemporalDynamicsFluvial2016; @teruiThreeEcologicalFactors2016], which collectively cover the entire Hokkaido island. For the Midwest, we assembled fish community data collected by the Iowa Department of Natural Resources, Illinois Environmental Protection Agency and Illinois Department of Natural Resources, Minnesota Pollution Control Agency, and Wisconsin Department of Natural Resources. We detailed procedures for initial data selection in **Supplementary Information**.

@@

We collected data for watershed characteristics (watershed area [km^2^], branching probability, and mean elevation [m]), climatic variables (annual mean air temperature [degree Celsius] and cumulative precipitation [mm]), and land use (the fractions of forest, urban, and agriculture, and dam density per watershed area [km^-2^]). We delineated watershed polygons and river lines (defined as $\ge$ 1 km^2^ in the watershed area) using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] with ArcMap 10.7 and SAGA 6.3.0 combined with the R package 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018]. Watershed area was estimated based on the delineated watershed polygons using functions in the R package 'sf'[@pebesmaSimpleFeaturesStandardized2018]. We estimated branching probability by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. We obtained mean elevations for each watershed polygon using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019]. We estimated the spatial averages of air temperature and cumulative precipitation at a 1-km resolution for each watershed, which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005] via the R package 'raster.' We calculated the fraction of forest, urban, and agriculture for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020]. We calculated dam density per watershed area (km^-2^) based on the Global Reservoir and Dam Database, Version 1 (GRanDv1)[@lehnerGlobalReservoirDam2011; @lehnerHighResolutionMappingWorld2011].

@@

We developed linear regression models to examine the effects of macro-scale drivers on diversity metrics. Log-transformed diversity metrics ($\alpha$, $\beta$ and $\gamma$ diversity) at watershed $i$, $log_{10}~y_i$, are assumed to follow a normal distribution as $log_{10}~y_i \sim Normal(\mu_i, \sigma^2)$. The parameter $\mu_i$ was related to linear predictors, including log-transformed watershed area (ecosystem size; $log_{10}~WA$), log-branching probability (ecosystem complexity; $log_{10}~P_b$), region (coded as Hokkaido = 0, Midwest = 1; $region$), air temperature, precipitation, and the fraction of forest land use (logit-transformed). Although our primary focus of this analysis is to assess the effects of watershed area and branching probability, we included climate and land use variables to control these effects statistically. We did not include the fractions of urban and agricultural land use in the statistical models because these variables are strongly correlated with the fraction of forest land use (**Figure S19**). Similarly, we did not use mean elevation in the statistical models because it showed a moderate correlation with air temperature (**Figure S19**). Correlations for other combinations of explanatory variables were weak to moderate.

@@

We constructed two models for each of the diversity metrics: global and region-specific models. The global model assumes that the effects of watershed area and branching probability are consistent across biogeographic regions (i.e., no interactive effects of the region with other linear predictors).

@@

# Acknowledgements
We acknowledge numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima for providing information on the Hokkaido Freshwater Fish Database HFish.

@@

# Data and code availability
Third parties provided data for empirical data analysis; references are provided in the main text and Supplementary Information where appropriate. R functions for the generation of branching networks and the metacommunity simulation are provided as the R package 'mcbrnet.' (available at https://github.com/aterui/mcbrnet). Codes for simulations, statistical analysis, figures, and tables are available at https://github.com/aterui/public-proj_stream-diversity.
